---
pagetitle: "jstats:One-way ANOVA"
title-prefix: ""
resources:
  - anova-follow-along.R
---

```{r}
#| label: obx-setup
#| include: false

# ---------------------------------------------------------------------------
# CONSOLE FACSIMILES -- render scaffolding, hidden (include: false). Same
# mechanism as quickstart.qmd and data.qmd: pane_facsimile.R defines the
# capture-once / display-twice helpers in their own dot-named environment
# (.obx). Each block below is one call whose code STRING is both printed into
# the page and executed once, so the shown code and the run code cannot drift.
# An unexpected error in any block HALTS the render (the publish gate); block
# 10's error is deliberate and is caught with expect_error = TRUE.
#
# Devices, by the OUTPUT-REGISTER rule: the page's prose and its "Reading this
# output" boxes READ the output of blocks 1-7, 9 and 10, so those panes render
# OPEN (.obx$show_console(fence = TRUE));
# block 11 repeats output already read and is the reader's own result to check,
# so it is a collapsed "What you should see" box (.obx$show_block()). Block 8's
# result is a plot, not Console output: its code is a plain fence and the
# figure is drawn by a chunk that runs the same code string.
#
# jstats is NOT pre-loaded, for data.qmd's reason: block 1 teaches
# library(jstats), and R prints the startup message on first attach only.
# joutput()/joptions() are left alone: a first-time reader runs on package
# defaults. The page writes nothing to disk, so there is no teardown.
#
# HOW-TO LAYOUT (S344). No top boxes: the shared material (before you start, how
# to use the pages, the finished script, reset your session) lives once on
# howtos.qmd, and this page points there. Under the first output: "If you're
# new to statistics", "Reading this output", "Why jstats does it this way".
# Under later outputs: one plain sentence of result in the main text, then a
# "Reading this output" box where there is something new to read.
# ---------------------------------------------------------------------------

options(width = 80)   # match a default 80-column console for table alignment
options(jstats.attach.noninteractive = TRUE)   # let block 1 show the real startup message
source("pane_facsimile.R")   # defines .obx (dot-named); turns the output color on
```

## One-way ANOVA

This page is a hands-on run-through of a one-way ANOVA with `jaov()`: which
version of the test to run, how to check the assumption that decides between
the two versions, which follow-up comparisons go with each, and what to put in
your write-up.

A one-way ANOVA (analysis of variance) compares the average of a numeric outcome
across the categories of one categorical variable — usually three or more
groups. The page assumes you have met the test in a course or a textbook, and
concentrates on running the test and reporting it.

This page follows the pattern of every How-To: numbered blocks that build one
script, with the finished script to download at the bottom.
[Using the How-To pages](howtos.qmd#using-the-how-to-pages) explains the
pattern, and how to reset your session if something goes wrong.

### 1. Load jstats and choose the data

We'll work on `clinic`, one of the two practice datasets that ship with the
package: 70 simulated clients in a study of four treatment conditions — a
control group, cognitive behavioral therapy (CBT), mindfulness training, and a
support group. `juse()` makes `clinic` the default data frame, so the commands
that follow can leave its name out.

```{r}
#| echo: false
#| results: asis
.obx$show_console(r"---(## ---- 1. Load jstats and choose the data -------------------------------

## Load jstats for this session, then make the clinic practice dataset
## the default, so the commands below can leave its name out
library(jstats)
juse(clinic))---", fence = TRUE,
  lead = "The command lines are shown; the standalone <code>##</code> comment lines aren't -- running a comment just echoes it back and does nothing.")
```

### 2. Run the standard ANOVA

The outcome is `Flourishing`, a wellbeing score from 0 to 100, and the grouping
variable is `Condition`. `jaov()` takes a formula, with the outcome on the left
of the tilde (`~`) and the grouping variable on the right.

```{r}
#| echo: false
#| results: asis
.obx$show_console(r"---(## ---- 2. Run the standard ANOVA ----------------------------------------

## Is there evidence that average flourishing differs across the four
## treatment conditions? The outcome goes left of the ~, the groups right
jaov(Flourishing ~ Condition))---", fence = TRUE)
```

The result is in the first row of the ANOVA table. The *p* is below .05, the
usual threshold, so the test is statistically significant: evidence that average
flourishing differs across the four conditions in the population from which the
sample was selected, and not only in the sample.

::: {.callout-tip collapse="true" title="If you're new to statistics"}
The four group means in the output differ; in real data they always do. That is
not the question. These 70 clients are a sample, and the question is about the
population from which the sample was selected: if the four conditions made no
difference in that population, could differences this size have turned up in the
sample simply through the luck of who was selected? The ANOVA is a test of
exactly that. (With data on a whole population, a census, there would be nothing
to test. The means you see would simply be the population's means.)
:::

::: {.callout-note collapse="true" title="Reading this output"}
- **The title** names the test. "One-Way ANOVA" is the standard version. The
  line under the title confirms which data frame was used, since the command
  left the data frame's name out.
- **Analysis N** is the number of cases the test used — all 70 here, because
  neither variable has missing values. When cases are set aside, a small Case
  Processing table takes this line's place and says how many and why (block 9
  has one).
- **Group Descriptives** gives each group's size, mean, and standard deviation,
  with a 95% confidence interval for the mean. Look here first: these are the
  differences in your sample.
- **The ANOVA table** is the test itself. Two numbers in its first row carry the
  result, *F* and *p*. The larger the *F* and the smaller the *p*, the stronger
  the evidence that the means differ in the population and not only in the
  sample. By the usual convention, a *p* below .05 counts as statistically
  significant. A significant result does not say the difference is large or
  important, and a non-significant result does not show the groups are the same
  in the population. The rest of the table, and the Eta-squared line under it,
  are covered in the books and in most statistics courses. You will need some of
  those numbers for a write-up; the last section of this page shows where to
  find them.
:::

::: {.callout-note collapse="true" title="Why jstats does it this way"}
One function runs both versions of this test, the standard ANOVA and Welch's
ANOVA, and one argument, `welch = TRUE`, switches between them. The same
argument does the same job in jstats's t-test, `jt()`. Both functions default to
the standard version, so a t-test and an ANOVA on the same two groups give the
same answer.

Base R arranges things differently. Its standard ANOVA is `aov()`; Welch's
version is a separate function, `oneway.test()`, with different output. Base R's
t-test, `t.test()`, runs Welch's version unless told otherwise, while `aov()`
runs the standard version, so on the same two groups the two functions give
different answers until you change a default. Levene's test and the Games-Howell
comparisons, both used later on this page, are not in base R at all. Add-on
packages fill those gaps, each with its own syntax, defaults, and output.

Base R also has to be told that the grouping variable is categorical.
`Condition` holds the codes 1 to 4 with labels attached, which is how data from
commercial statistical software usually arrive. Hand `Condition` to `aov()` as
it is, and R treats the codes as a measured quantity and runs a different
analysis, with no warning. `jaov()` reads the labels, makes the conversion for
you, and then runs base R's own `aov()` underneath — the same proven code, with
the setup done.
:::

### 3. The other version, Welch's ANOVA

jstats offers two versions of the test. The **standard ANOVA** you just ran
assumes that the scores are about equally spread out in every group — *equal
variances*, also called homogeneity of variance. **Welch's ANOVA** drops that
assumption.

The standard version is the default because it is the traditional one: the test
most courses and textbooks teach, the table commercial statistical software
prints first, and the one most readers of your work will expect unless you say
otherwise. Welch's version is one argument away: `welch = TRUE`.

```{r}
#| echo: false
#| results: asis
.obx$show_console(r"---(## ---- 3. The other version, Welch's ANOVA ------------------------------

## The same question, without assuming the groups are equally spread out
jaov(Flourishing ~ Condition, welch = TRUE))---", fence = TRUE)
```

Welch's ANOVA is significant too, so on these data the two versions lead to the
same conclusion. The two versions will not always agree, and the next two blocks
are about choosing between them.

::: {.callout-note collapse="true" title="Reading this output"}
The title now says Welch's. The test table is shorter than the standard one: it
shows *F*, two degrees-of-freedom values, and *p*, and the note under the table
says that sums of squares and mean squares do not apply to Welch's ANOVA. The
second degrees-of-freedom value carries a decimal, which is normal for this
test. *F* and *p* are read as in block 2. The Group Descriptives table is the
same as in block 2, because Welch's ANOVA changes how the means are *tested*,
not the means. The Eta-squared line has a note of its own, because the effect
size is still worked out the standard way.
:::

::: {.callout-note collapse="true" title="Going deeper (optional)"}
Some statisticians now recommend running Welch's version as a matter of routine,
on the grounds that it costs little when the variances are equal and protects
you when the variances are not. Others, and most courses, keep the standard test
unless there is a reason to leave it. Either position is defensible. What
matters is that you say which test you ran and why, and that you follow the
preference of your field, or of whoever will assess your work, where there is
one. The help page `?jaov` documents every option used on this page; Book 1
covers what the *F* test is doing and why unequal variances matter, in more
detail, with worked examples.
:::

### 4. Check equal variances with Levene's test

Whether the variances are equal is something you can check. Levene's test
compares the spread of the groups, and `levene = TRUE` adds Levene's test to the
output.

```{r}
#| echo: false
#| results: asis
.obx$show_console(r"---(## ---- 4. Check equal variances with Levene's test ----------------------

## Levene's test asks whether the groups differ in spread. It is off by
## default, so ask for it
jaov(Flourishing ~ Condition, levene = TRUE))---", fence = TRUE)
```

Levene's test is not significant here: its *p* is above .05. The test found no
evidence that the four conditions differ in spread in the population, so there
is no reason to doubt the standard ANOVA's assumption, and the standard ANOVA
from block 2 is the one to report.

Equal variances is the assumption most people have in mind when they talk about
checking an ANOVA, and it is the assumption that decides between the two
versions of the test. Both versions make two other assumptions as well: that the
cases are independent of one another, and that the scores within each group are
roughly normally distributed. The box plot in block 8 gives a first look at the
second of those.

::: {.callout-note collapse="true" title="Reading this output"}
The new table sits at the top, above the descriptives. Levene's test has an *F*
and a *p* of its own, and its *p* is read the other way around from most tests:
a *p* **above** .05 is the reassuring result. When Levene's test is not
significant, jstats adds nothing further, and the rest of the output is what
block 2 printed.

Glance at the SD column of Group Descriptives as well. Levene's test is a guide,
not a verdict: with small groups the test can miss a real difference in spread,
and with very large groups the test can flag a trivial one. Here the Control
group's SD is the largest of the four, though not by enough for Levene's test to
flag it; block 8 shows where that extra spread comes from.
:::

### 5. When Levene's test is significant

To see the other outcome of Levene's test we need a different variable.
`ScreenTime` is each client's daily screen time, in hours.

```{r}
#| echo: false
#| results: asis
.obx$show_console(r"---(## ---- 5. When Levene's test is significant -----------------------------

## A different outcome, daily screen time, whose spread does differ
## across the conditions. Watch for the note under the Levene table
jaov(ScreenTime ~ Condition, levene = TRUE))---", fence = TRUE)
```

This time Levene's test is significant: evidence that the four conditions differ
in spread. jstats adds a note under the Levene table, and because the four
groups are close to the same size, the note says the standard ANOVA remains
appropriate.

That gives the conventional rule for choosing between the two versions:

- Levene's test **not significant**: report the standard ANOVA.
- Levene's test **significant, group sizes about equal**: the standard ANOVA is
  usually still acceptable. Welch's ANOVA is the safer choice when the SDs are
  far apart, as some of the SDs are here. Say which version you used.
- Levene's test **significant, group sizes unequal**: use Welch's ANOVA.

The rule is a rule of thumb, not a guarantee. Whichever version you choose, say
what you did and why.

::: {.callout-note collapse="true" title="Reading this output"}
Levene's *p* is below .05, and the SD column shows why: the Control group's
screen time is bunched more tightly than the other groups'.

The note under the Levene table is there because unequal variances matter more
in some situations than in others. The standard ANOVA is least affected by
unequal variances when the groups are about the same size, and most affected
when the group sizes are unequal as well. So jstats checks the group sizes for
you. jstats treats the group sizes as approximately equal when the largest group
is no more than 1.5 times the size of the smallest, a rule of thumb. Here the
largest group has 21 clients and the smallest has 14, which is exactly 1.5. Had
the group sizes been further apart, the note would have said instead that the
unequal variances may affect the results, and suggested `welch = TRUE`.

The ANOVA itself is not significant in this output: the test gives no evidence
that average screen time differs across the conditions in the population.
Levene's test and the ANOVA answer different questions, one about spread and
one about means.
:::

### 6. Which groups differ, Tukey HSD

Back to `Flourishing`. A significant ANOVA is evidence that the group means are
not all the same in the population. A significant ANOVA does not say which
groups differ. That is the job of *post-hoc comparisons*: every pair of groups
compared, with the *p*-values adjusted so that making several comparisons at
once does not inflate the chance of a false alarm. `posthoc = TRUE` adds the
comparisons, and after the standard ANOVA the method is Tukey HSD (the letters
stand for "honestly significant difference").

```{r}
#| echo: false
#| results: asis
.obx$show_console(r"---(## ---- 6. Which groups differ, Tukey HSD --------------------------------

## Post-hoc comparisons test every pair of groups. After the standard
## ANOVA, the method is Tukey HSD
jaov(Flourishing ~ Condition, posthoc = TRUE))---", fence = TRUE)
```

Two of the six comparisons are significant: clients in CBT and clients in
Mindfulness each scored higher on flourishing than clients in the Control group.
None of the other four comparisons reaches significance, CBT against Mindfulness
included.

::: {.callout-note collapse="true" title="Reading this output"}
The new table is at the bottom. Each row is one pair of groups, and the row's
name says which way the subtraction runs: `CBT-Control` is the CBT mean minus
the Control mean, so a positive difference means CBT scored higher. Next comes
the 95% confidence interval for that difference, and then the adjusted *p*.
Because the adjustment is built in, you read each *p* against .05 as it stands.
The interval tells the same story: an interval that does not include zero goes
with a *p* below .05.

Four groups make six pairs, and five groups would make ten, so this table grows
quickly. If the table's lines wrap in your Console, drag the Console pane wider
and run the line again.
:::

### 7. Which groups differ after Welch's, Games-Howell

Tukey HSD carries the same equal-variances assumption as the standard ANOVA. If
you used Welch's ANOVA, the matching method is Games-Howell, which does not
assume equal variances either. You don't ask for Games-Howell by name:
`posthoc = TRUE` gives Tukey HSD after the standard ANOVA and Games-Howell after
Welch's ANOVA.

```{r}
#| echo: false
#| results: asis
.obx$show_console(r"---(## ---- 7. Which groups differ after Welch's, Games-Howell ---------------

## The same request after Welch's ANOVA gives Games-Howell comparisons,
## which do not assume equal variances either
jaov(Flourishing ~ Condition, welch = TRUE, posthoc = TRUE))---", fence = TRUE)
```

The conclusion is the same on these data: the same two comparisons are
significant, with somewhat larger *p*-values. Where the groups' variances really
do differ, the two methods can disagree, which is the reason to match the
post-hoc method to the test.

::: {.callout-note collapse="true" title="Reading this output"}
The table has the same layout as the Tukey HSD table, with one more column,
`df`. Games-Howell judges each pair on those two groups' own variances, so each
comparison has a degrees-of-freedom value of its own. You need the `df` column
only if your write-up reports it. The mean differences are the same as in the
Tukey HSD table, because they are simply differences between group means. The
intervals and, as a rule, the *p*-values are not the same.

One caution. Games-Howell has to estimate each group's variance from that group
alone, and with very small groups its *p*-values can come out too small. A rule
of thumb you will see quoted is at least six cases in every group.
:::

### 8. Look at the groups with jplot()

A table of means summarizes each group in a few numbers; a plot shows what the
numbers are summarizing. Save the ANOVA's result under a name with the
assignment operator, `<-`, and hand that name to `jplot()`.

```{r}
#| echo: false
#| results: asis
.b8 <- r"---(## ---- 8. Look at the groups with jplot() -------------------------------

## Save the result under a name, then hand it to jplot() for a box plot
## of each group. The plot appears in the Plots tab (lower-right pane)
Results <- jaov(Flourishing ~ Condition)
jplot(Results))---"
.fence <- strrep("`", 3)
cat(.fence, "r\n", .b8, "\n", .fence, "\n", sep = "")
```

The first line prints the same output as block 2 while saving the result as
`Results`. The second line draws a box plot of each group in the **Plots** tab
of the lower-right pane:

```{r}
#| label: anova-boxplot
#| echo: false
#| results: hide
#| message: false
#| fig-width: 7
#| fig-height: 4.5
#| fig-alt: "Four box plots of the flourishing score, one for each treatment condition, numbered 1 to 4 along the horizontal axis, with a diamond marking each group's mean. Condition 1 has one separate point far below its box, at a score of 0."
eval(parse(text = .b8))
```

One thing stands out: a single client in the Control group, far below everyone
else, with a flourishing score of 0. A score like that deserves a look at the
data before you rely on the result — is the score real, or a typing slip? (In
these practice data the score is simply very low.)

::: {.callout-note collapse="true" title="Reading this output"}
Each box spans the middle half of that group's scores, the heavy line across the
box is the median, and the diamond marks the mean — the number the ANOVA
compares. The lines above and below the box reach to the lowest and highest
scores that are not unusually far from the box, and any score beyond those lines
is drawn as a separate point. The horizontal axis shows each condition by its
code; the Group Descriptives table has the names (1 is Control, 2 CBT, 3
Mindfulness, 4 Support group).

Two things are worth looking for, because either can distort a mean and a
standard deviation. **Skew**: a median line well off the center of its box, or a
diamond some distance from the median line, means the scores are stretched out
in one direction. **Outliers**: the separate points. The Control client with a
score of 0 drags the Control group's mean down and stretches the Control group's
standard deviation; that one client is the main reason Control's SD is the
largest in the Group Descriptives table.
:::

### 9. A group with one case

Sometimes a category turns up with a single case in it: one respondent who
checked "Other," one site that recruited one participant. It is worth knowing
what each version of the test does then. No group in `clinic` is that small, so
this block makes one. The `subset =` argument keeps only some of the cases, for
this one call, and the rule written here keeps every client who is not in CBT,
plus one client who is.

```{r}
#| echo: false
#| results: asis
.obx$show_console(r"---(## ---- 9. A group with one case -----------------------------------------

## No group in clinic is that small, so make one for this call only:
## subset = keeps every client who is not in CBT (condition 2), plus one
## who is (!= means "is not equal to", and | means "or")
jaov(Flourishing ~ Condition, posthoc = TRUE,
     subset = Condition != 2 | ClientID == "C016"))---", fence = TRUE)
```

The standard ANOVA still runs when a group has one case. But the test learns
very little about that group, as the wide intervals on the CBT rows of the Tukey
HSD table show.

::: {.callout-note collapse="true" title="Reading this output"}
A Case Processing table has replaced the one-line N at the top. The table
reports that the `subset =` rule set aside 13 clients and left 57.

In Group Descriptives, the CBT row shows an N of 1 and a mean (that one client's
score) and then nothing. One case has no standard deviation, so the CBT row has
no confidence interval either. The standard ANOVA can still run because it
estimates the spread from the other three groups and applies that estimate to
all four. In the Tukey HSD table, the three rows that involve CBT have intervals
far wider than the other rows.
:::

### 10. One case under Welch's ANOVA

Welch's ANOVA cannot do the same. Welch's ANOVA needs each group's own variance,
and a single case does not have one. Run the same analysis as a Welch's ANOVA
and jstats stops — the one deliberate error on this page.

```{r}
#| echo: false
#| results: asis
.obx$show_console(r"---(## ---- 10. One case under Welch's ANOVA ---------------------------------

## The same cases as a Welch's ANOVA. It needs each group's own variance,
## and one case has none, so jstats stops and says so
jaov(Flourishing ~ Condition, welch = TRUE,  # an error, on purpose
     subset = Condition != 2 | ClientID == "C016"))---", fence = TRUE,
  expect_error = TRUE, expect_error_match = "only 1 case")
```

The error names the category, says what Welch's ANOVA requires (at least two
cases in every category), and gives the way out. Which way to take is a judgment
about your study, not about the software: keep the one-case group and use the
standard ANOVA, combine that group with a category it belongs with, or leave
that group out. Whichever you choose, say so in your write-up.

### 11. Everything in one call

Once you know the pieces, `full = TRUE` asks for all of them together: Levene's
test, the confidence intervals, the effect size, and the post-hoc comparisons.
Add `welch = TRUE` to the same call for the Welch route.

```{r}
#| echo: false
#| results: asis
.obx$show_block(r"---(## ---- 11. Everything in one call ---------------------------------------

## full = TRUE asks for Levene's test, the confidence intervals, the
## effect size, and the post-hoc comparisons together
jaov(Flourishing ~ Condition, full = TRUE))---",
  chrome = TRUE)
```

::: {.callout-tip collapse="true" title="Coming from SPSS, Stata, or SAS?"}
Everything on this page has a counterpart in the commercial programs, though
the pieces are arranged differently in each.

In **SPSS** the tests are all in one dialog: One-Way ANOVA, in the Analyze menu
under Compare Means (Compare Means and Proportions in recent versions). The
*Options* button holds the descriptives, the homogeneity of variance test
(Levene's), and Welch's test; the *Post Hoc* button holds Tukey and, in its
"Equal Variances Not Assumed" group, Games-Howell. SPSS labels the rows of its
ANOVA table Between Groups and Within Groups, where jstats names the grouping
variable and `Residual`. In recent versions SPSS's Levene table has several
rows; the test jstats runs is the classic one, built on each score's distance
from its own group's mean, which is the row SPSS labels "Based on Mean."

In **Stata**, `oneway Flourishing Condition, tabulate` runs the standard ANOVA
with the group means. The equal-variances test that command prints is
Bartlett's, a different test; Levene's comes from
`robvar Flourishing, by(Condition)`, where it is the statistic labeled W0. Tukey
comparisons are `pwmean Flourishing, over(Condition) mcompare(tukey) effects`.
Welch's ANOVA and Games-Howell are not part of those commands;
community-contributed commands exist for both.

In **SAS**, a `PROC GLM` step with `CLASS Condition;` and
`MODEL Flourishing = Condition;` runs the standard ANOVA, and one more
statement, `MEANS Condition / HOVTEST=LEVENE(TYPE=ABS) WELCH TUKEY;`, adds
Levene's test, Welch's ANOVA, and Tukey comparisons. The `TYPE=ABS` matters:
left to its default, SAS's Levene's test works from squared distances and gives
a different result from the one on this page. That `MEANS` statement has no
Games-Howell option.

One thing jstats adds is the pairing: ask for post-hoc comparisons and you get
the method that matches the test you ran.
:::


```{r}
#| label: prose-gates
#| include: false

# ---------------------------------------------------------------------------
# PROSE GATES AND WRITE-UP NUMBERS -- hidden.
# The prose on this page makes statements about what the output shows (which
# test is significant, which pairs differ, which note prints). The practice
# data are fixed, so the statements are true today; if clinic is ever rebuilt,
# or jaov() changes, a statement could go false while the page still renders.
# Each stopifnot() below is one such statement: a false one HALTS the render,
# the same publish gate the blocks have. The reporting examples under "What to
# write up" are built here from the returned objects, so their numbers are the
# ones the Console panes show.
# ---------------------------------------------------------------------------

.q <- function(expr) {            # run a jstats call quietly, keep what it printed
  out <- utils::capture.output(res <- suppressMessages(expr))
  attr(res, "printed") <- out
  res
}
.std <- .q(jaov(Flourishing ~ Condition, data = clinic, posthoc = TRUE, levene = TRUE))
.wel <- .q(jaov(Flourishing ~ Condition, data = clinic, welch = TRUE, posthoc = TRUE))
.scr <- .q(jaov(ScreenTime ~ Condition, data = clinic, levene = TRUE))
.one <- .q(jaov(Flourishing ~ Condition, data = clinic, posthoc = TRUE,
                subset = Condition != 2 | ClientID == "C016"))

# Levene's row, read from what jaov() printed (the returned object holds none)
.lev <- function(res) {
  out <- fansi::strip_sgr(attr(res, "printed"))
  i <- grep("Levene's Test for Homogeneity of Variance", out, fixed = TRUE)
  stopifnot(length(i) == 1L)
  v <- strsplit(trimws(out[i + 3L]), "\\s+")[[1]]
  stopifnot(length(v) == 4L)
  list(f = as.numeric(v[1]), df1 = v[2], df2 = v[3], p = v[4],
       note = paste(out[(i + 4L):(i + 7L)], collapse = " "))
}
.lev_std <- .lev(.std); .lev_scr <- .lev(.scr)
.pnum <- function(p) as.numeric(sub("^<", "", p))

.wd  <- .one$posthoc$upper - .one$posthoc$lower   # interval widths, block 9
.sig <- function(res) res$posthoc$comparison[res$posthoc$p < .05]
.two <- c("CBT-Control", "Mindfulness-Control")
.n   <- function(res) stats::setNames(res$descriptives$N, sub("^[0-9]+: ", "", res$descriptives$Group))

stopifnot(
  # block 2: all 70 cases, a significant F
  .std$n == 70, .std$p < .05,
  # block 3: Welch's agrees, and its second df carries a decimal
  .wel$p < .05, .wel$df2 != round(.wel$df2),
  # block 4: Levene's test not significant for Flourishing
  .pnum(.lev_std$p) > .05,
  # block 4: Control's SD is the largest of the four
  which.max(.std$descriptives$SD) == 1,
  # block 5: some SDs are "far apart" (largest more than twice the smallest)
  max(.scr$descriptives$SD) / min(.scr$descriptives$SD) > 2,
  # block 5: significant for ScreenTime; largest group 21, smallest 14, the
  # "approximately equal" note; Control the least spread; the ANOVA itself n.s.
  .pnum(.lev_scr$p) < .05, max(.n(.scr)) == 21, min(.n(.scr)) == 14,
  grepl("approximately", .lev_scr$note), grepl("standard test remains appropriate", .lev_scr$note),
  which.min(.scr$descriptives$SD) == 1, .scr$p > .05,
  # blocks 6 and 7: six comparisons; the same two pairs, and only those,
  # significant under both methods
  nrow(.std$posthoc) == 6, nrow(.wel$posthoc) == 6,
  setequal(.sig(.std), .two), setequal(.sig(.wel), .two),
  all(.std$posthoc$diff[.std$posthoc$comparison %in% .two] > 0),
  all(.wel$posthoc$p[.wel$posthoc$comparison %in% .two] >
      .std$posthoc$p[.std$posthoc$comparison %in% .two]),
  # block 8: one Control client at 0, the group's mean below its median
  sum(clinic$Flourishing == 0) == 1,
  as.numeric(clinic$Condition)[clinic$Flourishing == 0] == 1,
  mean(clinic$Flourishing[as.numeric(clinic$Condition) == 1]) <
    stats::median(clinic$Flourishing[as.numeric(clinic$Condition) == 1]),
  # block 8: without that client, Control's SD is in line with the others
  stats::sd(clinic$Flourishing[as.numeric(clinic$Condition) == 1 & clinic$Flourishing > 0]) < 13,
  # block 9: 13 clients set aside, 57 left, CBT down to one case, and the
  # three CBT rows of the Tukey table "far wider" than the other three
  .one$n == 57, .n(.one)[["CBT"]] == 1,
  min(.wd[grepl("CBT", .one$posthoc$comparison)]) >
    2 * max(.wd[!grepl("CBT", .one$posthoc$comparison)])
)

# ---- the reporting examples -------------------------------------------------
.f2  <- function(x) formatC(x, format = "f", digits = 2)
.p3  <- function(p) if (p < .001) "*p* < .001" else
  paste0("*p* = ", sub("^0", "", formatC(p, format = "f", digits = 3)))
.code <- c(Control = 1, CBT = 2, Mindfulness = 3)
.msd <- function(res, g) {         # from the data, so the rounding is exact
  y <- clinic$Flourishing[as.numeric(clinic$Condition) == .code[[g]]]
  paste0("*M* = ", .f2(mean(y)), ", *SD* = ", .f2(stats::sd(y)))
}
.pp  <- function(res, cmp) .p3(res$posthoc$p[res$posthoc$comparison == cmp])
.pps <- function(res) {           # "p = .017 for each", or the two in order
  a <- .pp(res, "CBT-Control"); b <- .pp(res, "Mindfulness-Control")
  if (identical(a, b)) paste0(a, " for each") else paste0(a, " and ", b, " respectively")
}
.groups <- function(res) paste0(
  "higher scores in the CBT group (", .msd(res, "CBT"), ") and the Mindfulness group (",
  .msd(res, "Mindfulness"), ") than in the Control group (", .msd(res, "Control"), "), ",
  .pps(res), "; no other pair of conditions differed significantly.")

.wu_std <- paste0(
  "Levene's test gave no indication of unequal variances, *F*(", .lev_std$df1, ", ",
  .lev_std$df2, ") = ", .f2(.lev_std$f), ", *p* = ", .lev_std$p,
  ", so a standard one-way ANOVA was used. Flourishing scores differed across the four ",
  "treatment conditions, *F*(", .std$df1, ", ", .std$df2, ") = ", .f2(.std$f), ", ",
  .p3(.std$p), ", η² = ", sub("^0", "", .f2(.std$eta_squared)),
  ". Tukey HSD comparisons showed ", .groups(.std))

.wu_wel <- paste0(
  "Welch's ANOVA, which does not assume equal variances, showed that flourishing scores ",
  "differed across the four treatment conditions, *F*(", .wel$df1, ", ",
  formatC(.wel$df2, format = "f", digits = 1), ") = ", .f2(.wel$f), ", ", .p3(.wel$p),
  ". Games-Howell comparisons showed ", .groups(.wel))
```


## What to write up

A results paragraph for a one-way ANOVA usually carries five things: which test
you ran, the *F* with its two degrees-of-freedom values, the *p*, an effect
size, and — when the test is significant — which groups differed, with their
means and standard deviations. If you checked equal variances, say what you
found: that check is the reason for your choice of test. The two examples below
follow the general pattern of APA style, common in the social sciences.
Conventions differ between fields and journals, so check what yours expects.

**The standard route**, from the output of blocks 4 and 6:

> `r .wu_std`

**The Welch route**, from the output of block 7:

> `r .wu_wel`

Every number comes from the output. The statistics are rounded to two decimal
places, a common convention; the *p*-values are left at three, and Welch's
second degrees-of-freedom value keeps the one decimal jstats prints. The two
values in parentheses after *F* are its degrees of freedom: in the standard
ANOVA table, the `df` of the first row and of the `Residual` row; in Welch's
table, `df1` and `df2`. η² is the Eta-squared line. *M* and *SD* come from Group
Descriptives, and the *p*-values of the comparisons come from the post-hoc
table. A paragraph like these usually sits beside a table of every group's mean
and standard deviation, so the reader can see the groups the sentences leave
out.

Two notes on the Welch example. In your own write-up, the first sentence would
also give your reason for choosing Welch's ANOVA, typically a significant
Levene's test with unequal group sizes. And the Welch example reports no effect
size: the eta-squared that jstats prints under Welch's table is worked out the
standard way, as the note under it says, so if you report that eta-squared, say
so.

## Take the whole script with you

That's the full run-through. If you'd like the entire script in one file to
keep, adjust, and reuse, download it here. Put your own data frame and variable
names in place of `clinic`, `Flourishing`, and `Condition`, drop the blocks you
don't need, and the script is the outline of your own analysis.

The file is the finished version of the script you built: the same code, in the
same order, with one difference made on purpose. The line in block 10 that
raises the deliberate error is commented out, with a note saying why, so the
file runs from top to bottom without stopping.
[Using the How-To pages](howtos.qmd#using-the-how-to-pages) says more about
finished scripts and how to run one.

<div style="margin: 1rem 0 2rem;">
<a href="anova-follow-along.R" download class="btn btn-primary">Download the full script (.R)</a>
</div>
