---
pagetitle: "jstats:Missing Values"
title-prefix: ""
resources:
  - missing-values-follow-along.R
---

```{r}
#| label: obx-setup
#| include: false

# ---------------------------------------------------------------------------
# EXPECTED-OUTPUT BOXES ("What you should see") -- render scaffolding, hidden
# (include: false). Same mechanism as data.qmd: pane_facsimile.R defines the
# capture-once / display-twice helpers in their own dot-named environment
# (.obx); each block below is one .obx$show_block() call whose code STRING is
# both printed into the page and executed once, so the shown code and the run
# code can never drift. An unexpected error in any block HALTS the render (the
# publish gate); block 3's error is deliberate and is caught with
# expect_error = TRUE.
#
# Differences from data.qmd's setup, each on purpose:
#   - jstats IS pre-loaded here. data.qmd leaves it unloaded because its block 1
#     teaches library(jstats) and wants the real startup message in the box;
#     this page's block 1 is a jload()/juse() pair, so a startup message would
#     be noise the reader did not ask for.
#   - No on-disk cleanup and no teardown. This page never saves a file: its
#     scope ends where the data leave R (that is data.qmd's territory), so there
#     is nothing on disk to clear and no real path can reach a box.
#   - The missing-value convention is put back to the package default before
#     any block runs. A first-time reader has no convention chosen; if the render
#     machine has one set (a .Rprofile, a previous page in the same session),
#     block 1's load message would phrase itself for that setting and block 2's
#     "none" line would have nothing to show. This is restoring the default a
#     reader arrives with, not setting a non-default.
# ---------------------------------------------------------------------------

options(width = 80)   # match a default 80-column console for table alignment
suppressPackageStartupMessages(library(jstats))
joptions(missing.convention = "none")
source("pane_facsimile.R")   # defines .obx (dot-named -> survives rm(list = ls()))
```

## Telling jstats about missing values

This page is a hands-on run-through of one small move that protects every
statistic you compute: telling jstats which numbers in a column are not really
data, and choosing the form jstats should keep that information in.

This page is about **coded** missing values — a −99 or a 999 standing in for a
refusal or a "not applicable." You need this page if the data you use, now or
later, may carry such codes, and most survey data do, whatever the format: a CSV
file (comma-separated values, the plainest spreadsheet format) brings the codes
with nothing to say what they mean, and a file from SPSS, Stata, or SAS usually
arrives with them declared, but not always. If your data have no missing values
at all, leave this page for later. Missing entries that are simply blank need no
declaring — they usually load as R's ordinary `NA` (short for "not available")
— but every analysis still
has to decide what to do with them, a topic of its own that the books cover in
full. No copy of SPSS, Stata, or SAS is needed to follow along; those names
appear only because the three ways of recording a missing value are named after
the programs that introduced them.

{{< include _before-you-start.qmd >}}

{{< include _how-to-use-this-page.qmd >}}

{{< include _new-to-r.qmd >}}

{{< include _reset-your-session.qmd >}}

### 1. Load the data, and see what jstats already noticed

We'll work on `clinic`, the deliberately messy practice dataset that ships with
the package. The first two lines are a pair you wouldn't normally write —
`clinic` is available by name the moment jstats is loaded, and `juse()` usually
comes once per session — but together they guarantee a fresh copy, whatever you
did to `clinic` on another page. Watch what jstats says as the data load, then
ask for the descriptives of `MoodRating`, a column that is meant to run from 1
to 10.

```{r}
#| echo: false
#| results: asis
.obx$show_block(r"---(## ---- 1. Load the data, and see what jstats already noticed ------------

## Start from a fresh copy of the clinic practice dataset
## You wouldn't normally write this pair: clinic is there by name as soon
## as jstats is loaded, and juse() usually comes once per session. The
## reload makes this page start clean whatever you did on another page
jload("clinic", overwrite = TRUE)
juse(clinic)

## MoodRating is meant to run from 1 to 10. Ask for its descriptives
jdesc(MoodRating))---",
  chrome = TRUE,
  lead = "The command lines are shown; the standalone <code>##</code> comment lines aren't -- running a comment just echoes it back and does nothing.")
```

Two things happened. As `clinic` loaded, jstats looked over its missing values
and reported on two groups of columns. Four of them — `Stress`, `SleepHours`,
`Medication`, and `Anxiety4` — arrived with their codes already *declared*, and
those codes are listed with labels such as "Refused". Two others, `MoodRating`
and `Anxiety2`, carry codes that jstats *suspects* stand for missing values but
that nobody has declared; for those it offers a ready-to-run example, naming
`MoodRating`. Then the descriptives showed why that second group matters: the
minimum of `MoodRating` is a large negative number and the mean has been dragged
well below the 1–10 range.
Those stand-in codes are being averaged in as if they were genuine ratings —
the usual state of affairs after a plain CSV or Excel import, where the file
carries no record of what the numbers mean.

::: {.callout-tip collapse="true" title="Coming from SPSS, Stata, or SAS?"}
Open that same CSV in SPSS and the codes sit in Data View looking like any
other number, Variable View shows no missing values defined, and Descriptives
averages them in without a word. The check jstats just ran on arrival is one
the commercial programs don't make: they trust the file. The four columns that
arrived already declared are the equivalent of a .sav whose Variable View was
filled in properly — jstats read those definitions and kept them.
:::

### 2. What a missing-value convention is

Before we fix `MoodRating`, a piece of background that the rest of the page
rests on. Every statistics program needs some way to mark a value as missing
while also remembering *why* — refused, don't know, not applicable. Base R was
not built with that need in mind: it has a single `NA` that says "missing" and
nothing more. The commercial programs came earlier and each solved the problem
in its own way, so three ways exist, one per program — and since much of the
data you will meet still arrive in SPSS, Stata, or SAS files, or from survey
platforms that export to them, one of those three is usually what your data
already use. jstats supports all three rather than inventing a fourth, and
leaves the choice of which to use to you. It calls that choice a
**missing-value convention**.

- **SPSS-style.** The code stays in the data as a visible number — a −99 is
  still a −99 in the column — and a note attached to the column says "treat
  −99 and −98 as missing." Every jstats function reads that note. The
  advantage is that nothing is hidden: you can see the codes in a frequency
  table and in the raw data.
- **Stata-style.** The code is replaced by a *marker*: `.a` for the first
  missing reason, `.b` for the second, and so on up to `.z`, each with its
  label. The markers are genuine missing values as far as R is concerned — which
  is why the package recommends this form; block 3 explains, and block 5 shows
  it live.
- **SAS-style.** The same idea as Stata, with uppercase markers (`.A` to `.Z`).

jstats will not pick one of these for you. The next block clears any convention
you may already have chosen, so that whatever you did on another page, this page
starts where a first-time reader starts — with no choice made. Block 3 then
shows what happens when you try to work without one.

```{r}
#| echo: false
#| results: asis
.obx$show_block(r"---(## ---- 2. What a missing-value convention is ----------------------------

## jstats can keep missing-value codes in any of three forms, one for each
## major program. It will not choose one for you. We start with no
## convention chosen -- the package default -- so you see what happens
## (If you've chosen one on another page or in your Project's .Rprofile,
## this line clears it for the current session only; restarting R brings
## your .Rprofile choice back)
joptions(missing.convention = "none"))---",
  chrome = TRUE)
```

::: {.callout-tip collapse="true" title="Coming from SPSS, Stata, or SAS?"}
If you have handled missing values in SPSS, Stata, or SAS, one of these three
will already be familiar — here it is under its own name. SPSS-style is
what MISSING VALUES in Variable View produces: the codes stay put and a
definition rides alongside. Stata-style is Stata's extended missing values
(`.a` through `.z`); SAS-style is SAS's (`.A` through `.Z`). The one new idea
is that jstats lets you choose *between* them for data that are now in R —
regardless of which program the file came from — and converts from one form to
another on request. Which form suits you is a question about how you work, not
a verdict on the programs: it depends on whether your data go back to one of
them, and on whether you also use base R alongside jstats.
:::

### 3. Declare, and meet the gate

`jdeclare_missing()` is the function that tells jstats which codes in a column
mean "missing," and it takes the labels at the same time. Run it now, exactly
as you will run it successfully in the next block — with no convention chosen,
it stops and asks.

```{r}
#| echo: false
#| results: asis
.obx$show_block(r"---(## ---- 3. Declare, and meet the gate ------------------------------------

## Declare -99 and -98 as missing on MoodRating, with labels
## With no convention chosen, jstats stops and asks instead of guessing
clinic <- jdeclare_missing(clinic, MoodRating,  # an error, on purpose
                           codes = c(Refused = -99, "Don't know" = -98)))---",
  chrome = TRUE, expect_error = TRUE,
  expect_error_match = "missing-value convention is selected")
```

This is the one place on the site where jstats refuses to do something it
plainly could do, and the refusal is deliberate. The three forms behave
differently once the data leave jstats — in base R, in other packages, in a
file bound for another program — so silently picking one would be choosing on
your behalf a consequence you had not seen. Instead the error lays out the three
choices, each with the one fact that separates it from the others, and puts its
recommendation first.

The recommendation is Stata-style, and the reason is the line under it in the
error: the markers behave as true NAs in base R. That holds for the SAS-style
markers just as much — both forms replace the code with a genuine missing value,
and the only difference between them is which letters they use. The real divide
is between those two and SPSS-style. Under SPSS-style, a −99 is still a −99,
and base R — which knows nothing about jstats's notes — will average it in and
hand you a plausible-looking wrong number. Under either marker
form the code has become a real missing value, so base R's `mean()` returns `NA`
instead: an answer you cannot mistake for a result, and one line fixes it. In
short, **SPSS-style fails silently in base R; the marker forms fail loudly, and
are fixable.** Stata-style is the one the recommendation names because it is the
more widely used of the two. Block 5 shows the contrast live, on the same column.

That still leaves room for judgment. If you write base R alongside jstats, or
paste in code an AI assistant wrote for you — which today will usually be base R
— choose Stata-style. If your data live in SPSS and go back there for a
collaborator, staying SPSS-style is a reasonable choice, and
jstats will read your declarations correctly either way. And notice the error's
last line: the same `joptions()` command, placed in your Project's `.Rprofile`,
makes the choice once and for all — block 7 says where that file lives.

::: {.callout-note collapse="true" title="Going deeper (optional)"}
Why do the programs differ at all? They made different trade-offs. SPSS keeps
the code visible as a number and attaches a definition beside it, which can suit
a survey codebook in the tradition that grew up around paper questionnaires,
where the code written on the form is itself the record of the refusal. Stata
and SAS fold the reason into the missing value itself, so that arithmetic can
never touch it by accident. Neither is wrong; they are answers to different
working habits. Book 1 covers what a missing-value code is, and what an
undeclared one does to any statistic that treats it as an ordinary number — the
mean is simply the easiest place to see it; Book 2 covers the three conventions,
how jstats converts between them, and what survives the trip to each file format
— in more detail, with worked examples. The help page `?jdeclare_missing`
documents the options.
:::

### 4. Choose, declare, check

Choose the recommended form, then run the same declaration again. It goes
through this time, and two checks show what changed: the descriptives, and a
frequency table. Then one more call — `joptions()` with nothing inside the
parentheses — shows your current settings, this new one among them.

```{r}
#| echo: false
#| results: asis
.obx$show_block(r"---(## ---- 4. Choose, declare, check ----------------------------------------

## Choose the Stata-style convention -- the one jstats itself recommends
## for anyone who also runs base R
joptions(missing.convention = "stata")

## Now the same declaration goes through
clinic <- jdeclare_missing(clinic, MoodRating,
                           codes = c(Refused = -99, "Don't know" = -98))

## The mean sits inside 1-10 again, and the codes count as missing
jdesc(MoodRating)

## A frequency table shows the codes set aside, with their labels
jfreq(MoodRating)

## With nothing in the parentheses, joptions() shows your current
## settings, including this one
joptions())---",
  chrome = TRUE)
```

Read the box from the top. Choosing the convention produced a note straight
away: `clinic` arrived with four SPSS-style columns, and jstats points out that
they no longer match your setting — the next block deals with that. The
declaration then reports what it did: each code became a marker (`.a`, `.b`)
carrying its label, and the note beneath reminds you that a call like this
changes `clinic` only because you assigned the result back with the assignment
operator, `<-` (the `modify = TRUE` form it mentions does the same job in
place — either is fine). It also points out, correctly, that `MoodRating` now
differs in form from the rest of `clinic`.

Then the payoff. The descriptives report a smaller non-missing count, a
minimum and maximum inside 1–10, and a mean that finally looks like a mood
rating. The line above each table still reads 70 cases, exactly as it did
before the declaration, and that is worth noticing: no rows were dropped. The
declared codes are missing within `MoodRating` rather than whole rows knocked
out of the dataset, and it is the non-missing count — 63 of 70 — that records
them. The frequency table is the clearest view of all: the
two codes sit in their own **Missing** section under their markers and labels,
counted in the totals and left out of the valid percentages. And the settings
listing confirms the convention is in place for the rest of your session.
(The row beneath it is always listed: those are the numbers jstats would use if
you ever converted your markers to the SPSS form.)

::: {.callout-tip collapse="true" title="Coming from SPSS, Stata, or SAS?"}
Every one of the commercial programs has a version of this move, and all three
get to the same analytical place: the codes stop counting as data, and no rows
are thrown away. They get there differently, though, and the difference is
worth knowing before you migrate. SPSS *declares*:
`MISSING VALUES MoodRating (-99, -98)`, with the matching VALUE LABELS, leaves
the −99s sitting in the cells and flags them — clear the declaration and they
are ordinary values again. Stata and SAS *replace*: the code is overwritten by
the program's own marker. In Stata that is
`mvdecode MoodRating, mv(-99=.a \ -98=.b)`, with `label define` supplying the
matching labels; in SAS it is a data-step assignment —
`if MoodRating = -99 then MoodRating = .A;`, and the same for `.B` — with
`PROC FORMAT` for the labels. Afterward the original number is not recoverable
from the data, though the labels still record what each marker meant. The
difference here is that the declaration you just made is not tied to any of
the three: it can be converted to either of the other forms without going back
to the raw data. Converting to the SPSS form puts numeric codes back, but which
numbers get used is set by a `joptions()` setting (`missing.convention.codes`),
because the original values are gone. The value labels, however, survive every
conversion unchanged.

The frequency table above is close to what SPSS's Frequencies, or Stata's
`tabulate` with missing values shown, would give you — a Valid section, a
Missing section, and valid percentages computed on the valid cases alone —
though the layouts are not identical. What has no counterpart in any of them is
the settings listing. Each program has only its own form, so there is nothing
to choose; in jstats there is, and `joptions()` is where you see what you chose.
:::

### 5. Bring the rest in line

`clinic` arrived with four columns already declared, SPSS-style — the common
case, since Qualtrics, LimeSurvey, and most public data archives hand out
SPSS-style files. jstats reads those declarations correctly under either
setting, so nothing is wrong yet. But base R does not read them, and this block
shows what that means before and after one line of conversion. `Stress` is one
of the four columns.

```{r}
#| echo: false
#| results: asis
.obx$show_block(r"---(## ---- 5. Bring the rest in line ----------------------------------------

## Stress arrived with its codes declared SPSS-style. Ask base R's mean()
## for its average, with and without na.rm = TRUE (which tells mean() to
## drop NAs). To base R the codes are ordinary numbers, so both go in
mean(clinic$Stress)  # base R -- the codes are averaged in
mean(clinic$Stress, na.rm = TRUE)  # na.rm has no NAs to drop yet

## One line converts every SPSS-style column in clinic to the Stata form
clinic <- jconvert(clinic, to = "stata")

## Now base R sees the missing values too: the mean is NA (loud, not wrong),
## and na.rm = TRUE gives the right answer
mean(clinic$Stress)
mean(clinic$Stress, na.rm = TRUE))---",
  chrome = TRUE)
```

There is the argument from block 3, live. Before the conversion, base R
returned a mean for `Stress` that looks perfectly reasonable and is wrong — the
−99s and −98s went into it — and `na.rm = TRUE` changed nothing, because to base
R those codes are not NAs; they are numbers. After `jconvert()` the same
column, holding the same information, gives `NA`: base R now *sees* the missing
values, refuses to average across them, and `na.rm = TRUE` does exactly what it
is for. The conversion report lists what moved, column by column. This is a
change to the data as they sit in R; what happens to declarations when you save
to a file and load it back is covered on
*[Loading, Saving, Converting Data](data.qmd)*.

::: {.callout-note collapse="true" title="What's the Base R equivalent?"}
The block *is* the base R equivalent, shown live. `mean()` and its `na.rm`
argument are base R's whole toolkit for missing values, and they only know one
kind: `NA`. A declared SPSS-style code is a jstats (and haven) idea layered on
top — base R never looks at the note, which is why the first mean is quietly
wrong. The conversion itself is haven's `tagged_na()` under the hood: the same
proven library the other pages mention, doing the fiddly part (one marker per
code, labels carried across, all four columns at once) so you don't. If you
would rather see the effect on one column than on all four, `jconvert()` takes
a `vars` argument; `?jconvert` has the details.
:::

### 6. Finding undeclared codes in your own data

`clinic` announced its problem columns at the door, because it arrived through
`jload()`. Your own data will not always be so polite — a data frame handed to
you by a colleague, or built by other code, comes with no message attached.
So it is worth knowing what an undeclared code looks like from the inside.
`jscreen()` screens the whole data frame; then a frequency table of the one
column still carrying the problem.

```{r}
#| echo: false
#| results: asis
.obx$show_block(r"---(## ---- 6. Finding undeclared codes in your own data ---------------------

## jstats named MoodRating and Anxiety2 as clinic loaded. Data that arrive
## some other way come with no message, so learn the symptoms: screen the
## whole data frame...
jscreen()

## ...then look at the one column still carrying the problem
jfreq(Anxiety2))---",
  chrome = TRUE)
```

Nothing in the screen says "undeclared code" in so many words — that message
came with the load. What the screen shows is the *symptoms*, and they are worth
learning because they are what you will see in a data frame that arrived some
other way. Look at `Anxiety2` against its neighbors. `Anxiety1`, `Anxiety4`, and
`Anxiety5` are classified as Likert items; `Anxiety2` is classified as plain
Numeric, because two stray values have stretched its range past anything a
five-point scale could hold. The same two values show up again in the outliers
column, where an item like this should have none. And in the missing-data
column `Anxiety2` reports nothing at all, while `Anxiety4` — the same kind of
item, with its codes declared — reports the missing cases it has. A frequency
table then settles it: the two codes sit in the table as rows of their own,
each with a valid percentage, counted as if they were answers.

There are two ways to deal with a column like that, and jstats's own messages
always suggest the first. **Declare** the codes, as you did for `MoodRating`:
nothing is deleted, the labels are kept, and the declaration can be revised or
removed later. Or **recode** them to `NA` with `jrecode()`: that also removes
them from every statistic, but it throws away the distinction between a refusal
and a don't-know, and there is no undo. Declaring is the careful default;
recoding is for when the reason genuinely does not matter.

So here is the exercise the page has been building toward. jstats told you at
the top of the page that `Anxiety2` carries the same two codes as `MoodRating`;
the screen and the frequency table have just shown you what that does. Declare
them yourself — the call is the one from block 4 with the column name changed —
then run the frequency table again and watch the codes move into the Missing
section. The finished script at the bottom of the page carries the answer, if
you want to check your work.

::: {.callout-note collapse="true" title="What's the Base R equivalent?"}
Base R has no way to *declare* a missing value; its only move is the recode.
The syntax is `clinic$Anxiety2[clinic$Anxiety2 %in% c(-99, -98)] <- NA`, and it
does exactly what the paragraph above warns about: both codes become the same
anonymous `NA`, so the difference between a refusal and a don't-know is gone,
and the original values are gone with it. It is the right tool when the reason for the missingness is irrelevant.
`jdeclare_missing()` exists because in most research it is not — the difference
between "refused" and "not applicable" is often the finding — and commercial
statistical software has always let you keep it.
:::

### 7. Make it permanent

The convention you chose is a session setting: it lasts until you restart R.
This page deliberately leaves it switched on when you move away — you chose it,
so it should still be there on the next page you visit, and one last look
confirms it is. To have jstats make the choice for you at
the start of every session in this Project, put the same
`joptions(missing.convention = "stata")` line in your Project's `.Rprofile` file;
the error in block 3 told you as much. *[Loading jstats
automatically](loading-jstats-automatically.qmd)* shows where that file lives and
how to create it.

```{r}
#| echo: false
#| results: asis
.obx$show_block(r"---(## ---- 7. Make it permanent ---------------------------------------------

## The convention you chose is a session setting: it lasts until you
## restart R. This page leaves it switched on -- one last look
joptions()

## To have it chosen automatically at the start of every session in this
## Project, put the same joptions(missing.convention = "stata") line in
## your Project's .Rprofile file. The Loading jstats automatically page
## shows where that file lives and how to create it)---",
  chrome = TRUE)
```

::: {.callout-note collapse="true" title="What's the point?"}
The download at the bottom of this page is the *finished* version of what you
built — and finished means something specific. A goal every experienced
programmer and statistician takes for granted, but new researchers are rarely
told: a completed script should run from beginning to end without errors. On
this page you met one error on purpose, because meeting it is how the choice
gets explained. The download shows what the end of that process looks like, and
it differs from the page blocks in two principled ways: the deliberate error in
block 3 is commented out, with a note saying why, and block 6 carries the answer
to the closing exercise — the `Anxiety2` declaration you were asked to write
yourself — so the file runs clean and complete. Nothing else changes: this page
saves no files, so there are no overwrite arguments to add, and the reload at
the top already guards against a second run.

To prove a script is finished, use the **Source** icon (to the right of Run in
the script toolbar): it runs the entire file, top to bottom, in one go. Source
shows the *output* without echoing each command back to the Console — that's
normal — and the small arrow beside Source offers "Source with Echo" if you'd
like the commands shown too. jstats functions print their own tables, so their
output always appears; but a bare base R line that merely displays a value —
`mean()` on its own — stays quiet under Source, because R only auto-prints
results when you run lines interactively.

The **What you should see** boxes under the blocks above reproduce that
Console — the blue command lines, the gray notes, the colored tables, and one
red error where it is meant to happen. They show the *code* being run, along
with any note written on the end of a command line, but not the standalone
`##` comment lines (running one just echoes it back and does nothing).

One thing this page leaves behind, deliberately: the missing-value convention
stays set for the rest of your session, because the whole point of choosing it
is to have it in place. Nothing else is changed and nothing is written to disk.
If you set a convention in your Project's `.Rprofile` and block 2 cleared it,
restarting R brings it back.

If you're keeping the file — and it's worth keeping, since finished scripts
travel — give it a proper home. In the **Files** tab of the lower-right pane,
click **New Folder** and type a name into the popup. "Scripts" is a good one,
and it will sit alongside whatever else your project folder holds.
:::

## Take the whole script with you

That's the full run-through. If you'd like the entire script in one file to
keep, adjust, and reuse, download it here:

The file is close to what you have just built: the same code, in the same
order, as though you had pasted every block above into one script. Two things
differ, each on purpose — the line that raises the deliberate error is
commented out, and the finished file includes the answer to block 6's exercise.
The **What's the point?** box under the last block explains why a finished
script is written that way.

<div style="margin: 1rem 0 2rem;">
<a href="missing-values-follow-along.R" download class="btn btn-primary">Download the full script (.R)</a>
</div>
